Formalization of Typed Lua

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: finished removal of void type

Second-level types are either tuple types or unions of tuple types, and represent sequences of values that appear as arguments, return values, and the left and right sides of
assignments. Tuple types are heterogeneous lists of first-class
types that end with a {\em variadic type}.
A variadic type $F{*}$ represents an unbounded sequence of values of type $F \cup \Nil$.

@@

Tuple types are covariant:
\[
\begin{array}{c}
\mylabel{S-PAIR}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
       \senv \vdash P_{1} \subtype P_{2}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\]

@@

Typed Lua preserves this behavior in assignment, but allows
both functions with strict arity, which use $\Void$, and
functions with flexible arity, which use other variadic types,
and can use $\Value{*}$ if they want to just ignore extra
arguments.

The types of expression lists always end in a variadic type
that is $\Nil{*}$ if the number of values the list produces
is statically known.

The other rules let assignment and functions with flexible
arity reflect the semantics of Lua in a type-safe way. We
need rule {\sc S-VARIADIC} because just having variadic
types be covariant would be too restrictive (for example,
$\Nil{*}$ would not be a subtype of $\Number{*}$):
\[
\begin{array}{c}
\mylabel{S-VARIADIC}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
      {\senv \vdash F_{1}{*} \subtype F_{2}{*}}
\end{array}
\]

Rules {\sc S-VAR-EXPAND} and {\sc S-VAR-DROP} state the
conditions when variadic tuples if different length are
compatible:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VAR-EXPAND}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
       \senv \vdash F_{1}{*} \subtype P_{2}}
      {\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VAR-DROP}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
       \senv \vdash P_{1} \subtype F_{2}{*}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
\end{array}
\end{array}
\]

>>>>

Second-level types are either tuple types or unions of tuple types, and represent sequences of values that appear as arguments, return values, and the left and right sides of
assignments. Tuple types are heterogeneous lists of first-class
types that end with a {\em variadic type}.
A variadic type $F{*}$ represents an unbounded sequence of values of type $F \cup \Nil$. Lua pads sequences with {\tt nil} to
have as many values as needed, and in Typed Lua sequences
where the number of values is statically known have a $\Nil{*}$
tail to reflect this semantics.

@@

Tuple types are covariant, so the subtyping rule for pair types is straightforward:
\[
\begin{array}{c}
\mylabel{S-PAIR}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
       \senv \vdash P_{1} \subtype P_{2}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\]

@@

Typed Lua preserves this behavior in assignment by synthesizing
tuple types that have a $\Value{*}$ tail for the left side
of an assignment. Non-variadic functions have a stricter
behavior: their parameter lists have a $\Nil{*}$ tail, so
a function can be passed any number of extra arguments
as long as they are literal {\tt nil}, but the type system will
reject extra arguments of any other type. Missing arguments
are also fine if the corresponding parameters have {\tt nil}
as a subtype: this is the Lua idiom for optional parameters.

Functions that want to keep the Lua behavior of just ignoring
any extra arguments can be variadic with type $\Value{*}$ as
the tail of their parameter tuple.

The remaining subtyping rules for second-level types
reflect this semantics. Rule {\sc S-VARIADIC} states
that variadic types are covariant (just having $F_1{*} \subtype F_2{*}$ iff $F_1 \subtype F_2$ would be too restrictive,
as $\Nil{*}$ would not be a subtype of $\Number{*}$):
\[
\begin{array}{c}
\mylabel{S-VARIADIC}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
      {\senv \vdash F_{1}{*} \subtype F_{2}{*}}
\end{array}
\]

Rules {\sc S-VAR-EXPAND} and {\sc S-VAR-DROP} state the
conditions when variadic tuples of different length are
compatible:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VAR-EXPAND}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
       \senv \vdash F_{1}{*} \subtype P_{2}}
      {\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VAR-DROP}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
       \senv \vdash P_{1} \subtype F_{2}{*}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
\end{array}
\end{array}
\]
